---
title: "Covid19Data"
author: "S. Scruggs"
date: "2024-03-05"
output:
  html_document: default
  pdf_document: default
---


 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(tidyverse)
library(lubridate)
```
## Covid 19 Data

I will start by reading in the data from the four main csv files. 
```{r get_jhu_data}
##Get current Data in the four files. 
# They all begin the same way. 
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <-c("time_series_covid19_confirmed_global.csv", 
               "time_series_covid19_deaths_global.csv", 
               "time_series_covid19_confirmed_US.csv", 
               "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)
```


Let's read in the data and see what we have.

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```


After looking at global_cases and global_deaths, I would like to tidy those datasets and put each variable (data, cases, deaths) in their own column. 

Also, I don't need Lat and Long for the analysis I am planning so I will get rid of those and rename Region and State to be more R friendly.

```{r tidy_global_data}
global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State','Country/Region',Lat, Long),names_to = "date", values_to = "cases") %>% select(-c(Lat,Long))

global_deaths <- global_deaths  %>% pivot_longer(cols = -c('Province/State', 'Country/Region',Lat, Long),names_to = "date", values_to = "deaths") %>% select(-c(Lat,Long))
  

global <- global_cases  %>% full_join(global_deaths) %>% rename(Country_Region = 'Country/Region', Province_State = 'Province/State') %>% mutate (date = mdy(date))

summary(global)

global <- global %>% filter(cases>0)
#This allows us to look at cases that are positive

summary(global)

US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat,Long_)) 

US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), names_to = 'date', values_to = 'deaths') %>% select(Admin2:deaths) %>% mutate(date = mdy(date)) %>% select(-c(Lat,Long_))


US <- US_cases %>% full_join(US_deaths)

global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>% left_join(uid, by = c("Province_State","Country_Region")) %>% select(-c(UID,FIPS)) %>% select(Province_State,Country_Region, date, cases, deaths, Population, Combined_Key)


```

## Visualizing
We are now going to start visualizing our data. 


```{r visualizing_data}
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum (deaths) , Population = sum (Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

US_totals <- US_by_state %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

tail(US_totals)

US_totals %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y= deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID19 in US", y = NULL)

state <- "Texas" 
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases"))+ geom_line(aes(y=deaths, color = "deaths")) + geom_point(aes(y= deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("COVID19 in ", state), y = NULL)

max(US_totals$date)
max(US_totals$deaths)


```

```{r analyzing_data}
US_by_state <- US_by_state %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))

US_totals %>% filter(cases > 0) %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y = new_deaths, color = "new_deaths")) + geom_point(aes(y= new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID19 in US", y = NULL)

state <- "Texas" 
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases"))+ geom_line(aes(y=new_deaths, color = "new_deaths")) + geom_point(aes(y= new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("COVID19 in ", state), y = NULL)

US_state_totals <- US_by_state %>%group_by(Province_State) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000* cases / population, deaths_per_thou = 1000* deaths / population) %>% filter(cases > 0, population > 0)

US_state_totals %>% slice_min(deaths_per_thou, n = 10) %>% select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>% slice_max(deaths_per_thou, n = 10) %>% select(deaths_per_thou, cases_per_thou, everything())


```
## Modeling Data


We are now going to model our data.

```{r modeling_data}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

summary(mod)

US_state_totals %>% slice_min(cases_per_thou)

US_state_totals %>% slice_max(cases_per_thou)

#x_grid <- seq(1,151)

#new_df <- tibble(cases_per_thou = x_grid)

US_state_totals %>% mutate(pred = predict(mod))

US_total_w_pred <- US_state_totals %>% mutate(pred=predict(mod))

US_total_w_pred

US_total_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue")+ geom_point(aes(x = cases_per_thou, y = pred), color= "red")


```

## Bias

It is important to always understand bias in data science, so we must understand where bias could come from here. Many people perhaps did not believe in COVID so they did not report it. This could drastically change our number of cases. Personal bias could also change the data and how we clean/organize it. 